Ensemble Stability of Biological Networks

### 2014-10-13 tex/introduction.tex
refactor introduction

%!TEX root = ../paper.tex
The traditional approach taken in the study of chemical reaction, gene-regulatory, population, and ecosystem networks is to consider a particular example, derive a system of differential equations to model that example, try to fit the model to data and adjust the modeling assumptions along with parameter values until a good fit is achieved \cite{Meyer2014}. Recent work has demonstrated that as a result of sloppiness in the dependence of qualitative dynamic phenomena on the geometry of parameter space that this approach allows for a large variety of models to fit the data \cite{Brown2003,Gutenkunst2007,Daniels2008a,Machta2013,Hines2014,Prabakaran2014,Tonsing2014}. In the face of uncertainty about the structure of such biological networks, to model the components under consideration as randomly interlinked becomes a reasonable approximation. This approach enables one to gain insight into what dynamical phenomena are possible to observe within a given class of dynamical systems, which is necessary to understand in order to determine whether or not a particular dynamical phenomenon should be regarded as unique or generic in the development and investigation of models applied to particular systems \cite{Gunawardena2013,Gunawardena2014}.

Indeed, investigating generic properties of a large class of dynamical systems was the approach taken by May in models of ecosystem dynamics \cite{Gardner1970,May1972}. The class of dynamical systems studied by May is so general that to restrict its applicability to ecosystem dynamics is certainly not necessary and perhaps even inefficient with respect to the goal of understanding how biological networks operate across the various relevant levels of organization. For example, the study of chemical reaction, population, and gene-regulatory networks all utilize essentially equivalent mathematical structure as that discussed by May in the context of ecosystem dynamics \cite{RossCr2003,Alon2006,Palsson2006,HamidBolouri2008,Palsson2011a,Voit2012,Sauro2012}. Developing unified mathematical descriptions of all of these is one of the paramount goals of systems biology. This incredibly generic applicability of gaining a better understanding of the class of models investigated by May demonstrates the unequivocal value of deeper investigation.  However, work attempting to continue the development of the so-called May-Wigner stability theorem revealed that May's conjectured stability criterion was not as easy to demonstrate as was initially believed \cite{Cohen1984,May1972a,Radius2014}.

>>>>

%!TEX root = ../paper.tex
The traditional approach taken in the study of chemical reaction, gene-regulatory, population, and ecosystem networks is to consider a particular example, derive a system of differential equations to model that example, try to fit the model to data and adjust the modeling assumptions along with parameter values until a good fit is achieved \cite{Meyer2014}. All of these models utilize essentially equivalent mathematical structure \ref{fig:biomodelexamples} \cite{RossCr2003,Alon2006,Palsson2006,HamidBolouri2008,Palsson2011a,Voit2012,Sauro2012}. Developing unified mathematical descriptions of all of these is one of the paramount goals of systems biology.

Recent work has demonstrated that as a result of \emph{sloppiness} in the dependence of qualitative dynamic phenomena on the geometry of parameter space that this approach allows for a large variety of models to fit the data \cite{Brown2003,Gutenkunst2007,Daniels2008a,Machta2013,Hines2014,Prabakaran2014,Tonsing2014}. In the face of uncertainty about the structure of such biological networks, to model the components under consideration as randomly interlinked becomes a reasonable approximation. This approach enables one to gain insight into what dynamical phenomena are possible to observe within a given class of dynamical systems, which is necessary to understand in order to determine whether or not a particular dynamical phenomenon should be regarded as unique or generic in the development and investigation of models applied to particular systems \cite{Gunawardena2013,Gunawardena2014}.

Indeed, investigating generic properties of a large class of dynamical systems was the approach taken by May in models of ecosystem dynamics \cite{Gardner1970,May1972}. The class of dynamical systems studied by May is not restricted to ecosystem dynamics and encompasses, among others, the dynamics of all of the biological networks represented in \ref{fig:biomodelexamples}. In particular, May conjectured on the basis of results from random matrix theory what eventually came to be referred to as the May-Wigner stability theorem \cite{Cohen1984,May1972a,Radius2014}. This relationship implies a relationship between a topological property, system connectivity, and a dynamical property, stability.

% The generic applicability of gaining a better understanding of the class of models investigated by May demonstrates the unequivocal value of deeper investigation.  However, work attempting to continue the development of the so-called May-Wigner stability theorem revealed that May's conjectured stability criterion was not as easy to demonstrate as was initially believed.
